---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r echo = FALSE}
# Avoid hard-coding the package name.
pkg_name = read.dcf('DESCRIPTION')[, 'Package']
pkg = paste0('<span class="pkg">', pkg_name, '</span>')

registerS3method(
  'knit_print', 'data.frame',
  \(x, options, ...) knitr::asis_output(paste(knitr::kable(x, ...), collapse = '\n')),
  asNamespace('knitr')
)
```


# `r pkg` <img src="man/figures/logo.svg" align="right" alt="" width="120" />

<code>`r pkg_name`::alias()</code> allows creating aliases for other R names or arbitrarily complex R expressions. Accessing the alias acts as-if the aliased expression were invoked instead, and continuously reflects the current value of that expression: updates to the original expression will be reflected in the alias; and updates to the alias will automatically be reflected in the original expression.


## Installation

You can install the development version of `r pkg` from GitHub:

```r
pak::pak('klmr/aka')
```


## Example

The simplest case aliases another name:

```{r echo = FALSE, message = FALSE}
library(pkg_name, character.only = TRUE)
```

```{r echo = FALSE, results = 'asis'}
writeLines(paste0('```r\nlibrary(', pkg_name, ')\n```'))
```

```{r example}
x = 'hello'
alias(ax = x)
ax

x = 'world'
ax

ax = 'goodbye'
x
```

As we can see, updates to the original names are reflected in the alias, and updates to the alias are reflected in the original name.

As an alternative syntax, we can use an “assignment-like” form:

```{r example-assign}
ax %=&% x
```

This form is strictly equivalent to `alias(ax = x)`.

Furthermore, aliases can be created for complex expressions:

```{r example-expr}
mercedes %=&% mtcars[grep('^Merc ', rownames(mtcars)), ]
mercedes
```

… and we can even update parts of an aliased expression to modify parts of the original, underlying objects:

```{r}
# Set all Mercedes engine types to V-shaped
mercedes$vs = 0

mtcars[8 : 14, 'vs', drop = FALSE]
```

However, be careful when assigning to an alias of an object in a parent environment:

```{r}
e = attach(new.env())
e$y = 'hello'

alias(ay = y)
```

This will seem to work:

```{r}
ay
```

… because `y` is found in the attached *parent environment* after it was not found in the current scope; however, the following will create a *new* variable `y` in the current scope, and leave `e$y` untouched:

```{r}
ay = 'world'
e$y
y
```

To prevent this: pass `env_expr` when defining the alias:

```{r eval = FALSE}
alias(ay = y, expr_env = e)
```

```{r echo = FALSE}
detach()
```


## Why?!

`r pkg` implements an approach to [reactive programming][] that is complementary to [Shiny’s reactive programming API][shiny-reactive]. Superficially, the mechanism resembles [references in C++][cpp-references]. However, under the hood they are fundamentally different. In particular, `r pkg` aliases are currently implemented as [active bindings][], whereas C++ references are direct aliases that are either replaced with the actual value by the compiler or converted to pointers.

This package purely provides syntactic sugar for active bindings, it does not add any *functionality* over manually calling `makeActiveBinding()` (the aliases created by `r pkg` *are* active bindings). It is also superficially similar to the <span class="pkg">[pointr]</span> package. Unlike the latter, the API of `r pkg` uses proper R expressions instead of being [stringly typed][], and its usage and implementation are conceptually more straightforward and idiomatic. To allow programming on the language, `r pkg` supports `bquote()`’s macro interpolation syntax rather than requiring string manipulation. It also allows explicitly controlling the evaluating and defining environments of the alias.

Truth be told, for now I have not yet identified a compelling use-case for this package; it mainly exists to explore the concepts, and to provide a playground. Another purpose is to act as a tutorial for the implementation of a simple but powerful non-standard evaluation API. If you’ve found an interesting use-case, please reach out: I am genuinely curious!


[reactive programming]: https://en.wikipedia.org/wiki/Reactive_programming
[shiny-reactive]: https://mastering-shiny.org/basic-reactivity.html
[cpp-references]: https://en.wikipedia.org/wiki/Reference_(C%2B%2B)
[active bindings]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/bindenv.html
[pointr]: https://cran.r-project.org/package=pointr
[stringly typed]: https://wiki.c2.com/?StringlyTyped
